---
title: "Data Exploration"
date: "12/7/2019"
output: pdf_document
---

```{r}
library(tidyverse)
library(magrittr)
library(janitor)
```
```{r}
##data wrangling ----
df <- read_csv("ipums_time_used.csv") %>% 
  clean_names() %>%
  filter(!is.na(wb_resp)) %>%
  rename(
    residence = metro, 
    education = educ, 
    employment_status = empstat,
    wellbeing_response = wb_resp,
    hourly_wage = hourwage, 
    hours_work_wk = uhrsworkt) %>% 
  select(-wt06, -pernum, -lineno, -caseid) %>%
  mutate(sex = case_when(
           sex == 1 ~ "Male",
           sex == 2 ~ "Female"),
         employment_status = case_when(
           employment_status == 1 ~ "Employed",
           employment_status == 2 ~ "Employed", 
           employment_status == 3 ~ "Unemployed",
           employment_status == 4 ~ "Unemployed",
           employment_status == 5 ~ "Not in labor force"),
         education = case_when(education < 20 ~ "BelowHS",
                               education == 20 ~ "High School",
                               education == 21 | education == 30 ~ "Some College",
                               education == 31 | education == 32 ~ "Associate Degree",
                               education == 40 ~ "Bachelor's Degree",
                               education == 41 ~ "Master's Degree",
                               education == 42 ~ "Professional Degree",
                               education == 43 ~ "Doctoral Degree"),
         race = case_when(
           race == 100 ~ "White",
           race == 110 ~ "Black",
           race == 120 ~ "American Indian",
           race == 131 ~ "Asian",
           race == 132 ~ "Pacific Islander", 
           race == 200 | race == 210 | race == 211 | race == 212 | race == 300 | race == 400 ~ "Black-Mixed", 
           race == 201 | race == 202 | race == 203 | race == 310 | race == 320 ~ "White-Mixed",
           race == 220 | race == 230 ~ "Other-Mixed"),
         residence = case_when (
           residence == 1 ~ "Metropolitan: Central City",
           residence == 2|residence == 3 ~ "Metropolitan: Others", 
           residence == 4 ~ "Nonmetropolitan")) %>% 
  filter (!is.na(residence))
```

##Questions: 

1. Does location effect happiness of a race? 

2. Are people happier when they make more money? 

##Hypotheses: 

1. Location and Race does not have an effect on someone's happiness 
$\beta{_0} = 0$
$\beta{_A} \neq 0$ 

2. Someone's hourly wage does not have an effect on someone's happpiness 
$\beta{_0} = 0$
$\beta{_A} \neq 0$ 

##Model 1: 

```{r}
race_location <- glm(wellbeing_response ~ race + residence, data = df)
summary(race_location)
```

##Model 1 Analysis:

##Model 2: 

```{r}
wage_logm <- glm(wellbeing_response ~ hourly_wage, data = df, family = binomial)
summary(wage_logm)
```

logit($\pi$) = 2.709e+00 - -1.225e-04 * hourly wage 

##Model 2 Analysis:

##Model 3: 

##Model 3 Analysis:

##Model 4: 

##Model 4 Analysis:
 


