---
title: "Data Exploration"
date: "12/7/2019"
output: pdf_document
---

```{r}
library(tidyverse)
library(magrittr)
library(janitor)
```

```{r}
##data wrangling ----
df <- read_csv("ipums_time_used.csv") %>% 
  clean_names() %>%
  filter(!is.na(wb_resp)) %>%
  rename(
    residence = metro, 
    education = educ, 
    employment_status = empstat,
    wellbeing_response = wb_resp,
    hourly_wage = hourwage, 
    hours_work_wk = uhrsworkt) %>% 
  select(-wt06, -pernum, -lineno, -caseid) %>%
  mutate(sex = case_when(
           sex == 1 ~ "Male",
           sex == 2 ~ "Female"),
         employment_status = case_when(
           employment_status == 1 ~ "Employed",
           employment_status == 2 ~ "Employed", 
           employment_status == 3 ~ "Unemployed",
           employment_status == 4 ~ "Unemployed",
           employment_status == 5 ~ "Not in labor force"),
         education = case_when(education < 20 ~ "BelowHS",
                               education == 20 | education == 21 | education == 30  ~ "High School",
                               education == 31 | education == 32 ~ "Associate Degree",
                               education == 40 ~ "Bachelor's Degree",
                               education == 41| education == 43 ~ "Master's Degree and above"),
         race = case_when(
           race == 100 ~ "White",
           race == 110 ~ "Black",
           race == 120 ~ "American Indian",
           race == 131 ~ "Asian",
           race == 132 ~ "Pacific Islander", 
           race == 200 | race == 210 | race == 211 | race == 212 | race == 300 | race == 400|
           race == 201 | race == 202 | race == 203 | race == 310 | race == 320 |
           race == 220 | race == 230 ~ "Mixed"),
         residence = case_when (
           residence == 1 ~ "Metropolitan: Central City",
           residence == 2|residence == 3 ~ "Metropolitan: Others", 
           residence == 4 ~ "Nonmetropolitan"),
         sex = as.factor(sex),
         residence = as.factor(residence),
         education = as.factor(education),
         employment_status = as.factor(employment_status),
         race = as.factor(race)) %>% 
  filter (!is.na(residence))
```

```{r}
lapply(df, class)
```

##Todo 

1. Pick model 
2. Run tests 
- check assumptions of linearity, normality, homoscedacity by looking at residual plots - use the overall F-test 
- interpret the R^2 value 
- decide if each predictor is needed with t-tests 
- calculate and intepret the estimated odds ratio 
- calculate the 95% CI for the odds ratio 
- calculate the probability of... 
3. Summarize results 


##Questions: 

1. Does location effect happiness of a race? 

2. Are people happier when they make more money? 

##Hypotheses: 
We hypothesize that attributes that are considered successful in society will have a positive effect on someone's wellbeing. 

1. Location and Race does not have an effect on someone's happiness 
$H_o: \beta{_1} = 0$
$H_a:\beta{_1} \neq 0$ 

2. Someone's hourly wage does not have an effect on someone's happpiness 
$H_o: \beta{_1} = 0$
$H_a:\beta{_1} \neq 0$ 

##Models  
```{r}
df$employment_status <- relevel(df$employment_status, ref = "Not in labor force")
```

```{r}
#Does race have an effect on someone's wellbeing? 
race <- glm(wellbeing_response ~ race, data = df, family = binomial(logit))
summary(race)
```

```{r}
#Does hourly wage have an effect on someone's wellbeing? 
wage_logm <- glm(wellbeing_response ~ hourly_wage, data = df, family = binomial(logit))
summary(wage_logm)
```

logit($\pi$) = 2.709e+00 - -1.225e-04 * hourly wage 

```{r}
exp(cbind(OR = coef(wage_logm), confint(wage_logm)))
```


```{r}
#Does age status has an effect on someone's happiness? 
age_logm <- glm(wellbeing_response ~ age, data = df, family = binomial(logit))
summary(age_logm)
```

```{r}
exp(cbind(OR = coef(age_logm), confint(age_logm)))
```

```{r}
#Does education level have an effect on someone's wellbeing? 
log_edu <- glm(wellbeing_response ~ education + employment_status + age + race + residence + hours_work_wk, data = df, family = binomial(logit))
summary(log_edu)
```

```{r}
#Does employment status has an effect on someone's happiness? 
emp_logm <- glm(wellbeing_response ~ employment_status, data = df, family = binomial(logit))
summary(emp_logm)
```

```{r}
exp(cbind(OR = coef(emp_logm), confint(emp_logm)))
```

```{r}
work_logm <- glm(wellbeing_response ~ hourly_wage + hours_work_wk + residence, data = df) 
summary(work_logm)
```

##Results so far:
- **Race**: Only asian is significant
- **Hourly wage**: significant on its own but not with number of hours and residence
- **number of hours worked**: significant
- **Area of residence**: significant
- **employment status**: significant
- **Age**: significant
- **Education**: only one category is significant 


```{r}
full_model <- glm(wellbeing_response ~ education + residence + employment_status + age, data = df, family = binomial(logit))
summary(full_model)
```

```{r}
## dataset with only employed people ----
employed_df <- df %>%
  filter (hours_work_wk != 9999, hourly_wage != 999.99) %>%
  select(wellbeing_response, employment_status, hours_work_wk, hourly_wage)
```

```{r}
employed_logm <- glm(wellbeing_response ~ hours_work_wk + hourly_wage, dat = employed_df, family = binomial(logit))
summary(employed_logm)
```

